### 2019-10-28 main_text/main_text.tex
Fix funding acknowledgement
Edits by KL using RP & MM feedback

% Title must be 250 characters or less.
\begin{flushleft}
	{\Large
		\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
	\\
	Kathrin Laxhauber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Dept???, ETHZ, Zurich, State???, Switzerland\todo{MJM: Add dept \& complete affiliation}
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Membrane transporters carry important molecules through the cell membrane
and, from a resource standpoint, should only be produced when necessary. The
expression of membrane transporters in metabolic pathways is often
upregulated by the transporter substrate. In \emph{E. coli}, such systems
include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
which encode for lactose, arabinose, and xylose transporters, respectively.
As a case study, we build a physical model of the \emph{xapABR} genetic
circuit, which features a regulatory feedback loop through membrane
transport (positive feedback) and enzymatic degradation (negative feedback)
of an inducer. Dynamical systems analysis and stochastic simulations show
that the membrane transport makes the model system bistable in certain
parameter regimes. Thus, it serves as a genetic “on-off” switch, enabling
the system to only produce a set of metabolic enzymes when the corresponding
metabolite is present in large amounts. We find that the negative feedback
from the degradation enzyme does not significantly disturb the positive
feedback from the membrane transporter. We investigate hysteresis in the
switching and discuss the role of cooperativity and multiple binding sites
in the model circuit. Altogether, this work explores how a stable genetic
switch for a set of enzymes is obtained from transcriptional auto-activation
of a membrane transporter through its substrate.

@@

\section*{Introduction}
Genetic regulatory circuits are fundamental building blocks of functioning
cells and organisms. One abundant class of these circuits are genetic
switches. Although their construction and function may differ,
their common feature is bistability: their output gene expression
will flow to and remain at one of two steady-state levels.
The distribution of
gene expression in a cell culture can then be bimodal. This is not to be
confused with mere stochastic bimodality, where the system is not stable,
and the gene expression in each cell can fluctuate between the two levels.

@@

\begin{figure}%[h!]
	\centering
	\includegraphics{media/Circuits.eps}
	\caption{{\bf A schematic of different genetic switches.}
	(A) and (B) show the two most well-known genetic switches:
	(A) two mutual repressors and (B) a self-activating gene.
	In (C), a very much simplified version of the
	circuit that we investigate in this paper can be seen, where the
	similarity to the switch in (B) becomes obvious. A complete version of
	the model circuit can be found in Fig~\ref{fig2:model}.}
	\label{fig1:switches}
\end{figure}

Through physical and mathematical modeling, we investigate a more complex
but common switch system where the bistability is due, as we will show, to a
membrane transport protein. Its key feature is the indirect activation of
the transporter gene by the transporter substrate, leading to positive
feedback similar to the aforementioned ``on-off'' switch. An example for
such an architecture is the \emph{lac} operon, where lactose indirectly
activates the expression of lactose permease. Other examples in \emph{E.
coli} include the \emph{araFGH} and \emph{xylFGH} operons, which contain genes for
arabinose and xylose transporters, respectively. For \emph{lac} and
\emph{araFGH}, bistability has indeed been observed and attributed to such a
positive feedback
loop~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
A eukaryotic example is the glucose transporter GLUT-2 in liver and
\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
complex than the following analysis.

It is quite conceivable that this auto-activation process is common to many
substances that a cell would want to accumulate. Such a switch enables the
cell to sense and respond to its environment: if the substrate enters the
cell, it activates the production of membrane transporters. The cell then
starts accumulating the substrate, thereby ``testing'' the substrate's
presence in the extracellular environment. If there is enough, the
expression stabilizes at an ``on'' point and the cell has, in a short-term
sense, adapted. When there is not enough substrate anymore, the operon, which
often encodes for a whole set of enzymes for this one metabolite, switches
``off'' again. Such a mechanism could be involved in various cases of
short-term adaptation.

A key element of this mechanism is the presence of a transcription factor
which is always expressed at a low level (often at copy numbers
of order $\sim10$) and which binds to the transporter substrate.
This is resource efficient for the cell, as this low copy number
transcription factor acts as an ``always on'' sensor to detect
the substrate, allowing high copy numbers of the membrane transporter
and its attendant operon to be expressed only when
their substrate is actually present. The transcription factors
LacI, AraC, and XylR all appear to fill this
role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
\todo{MJM: do we have a cite for xylR too?}

For our modeling, we focus on the \emph{xapABR} genetic circuit from
\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
complex. Instead of lactose, its purpose is to make use of the nucleoside
xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
circuit is made up of two operons: one that encodes for XapR and another
that encodes for XapA and XapB. XapR appears to be a transcription factor
that is induced by xanthosine and activates the \emph{xapAB} promoter,
in close analogy to LacI, AraC, and XylR.
The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
but the promoter architecture and function is not yet fully understood. The
transcription of \emph{xapR} seems to be constitutive and not
auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
factors suggests that XapR appears in dimers where one dimer can bind two
xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
nucleoside phosphorylase that degrades xanthosine into components (ribose
and xanthine) that can be fed into metabolic
pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

@@

\section*{Model}
\subsection*{Step by step modeling of the system}
In this section, we present our model of the \emph{xapAB} genetic circuit.
Fig~\ref{fig2:model} shows an overview of this model.
The qualitative picture of the circuit switching its state is as follows:
\begin{itemize}
	\item In the initial absense of XapB, small amounts of xanthosine are
		nonspecifically transported into and out of the cell (discussed in more detail below).
	\todo{MJM: ``nonspecific transport'' is a bit misleading:
	it's actually weakly specific transporters, but the precise mechanism is unimportant
	\& I think it better to emphasize generality of the model. Does this seem fine?}
	\item XapR is induced by xanthosine.
	\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
	\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
	\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
	\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
\end{itemize}
Because xanthosine induces the transcription factor XapR, we have positive
and negative feedback loops due to XapB and XapA, respectively. The remainder
of this subsection discusses each of the above steps in detail, leading us
to a set of two coupled ODEs. More in-depth explanations can be found
in~\nameref{S1_Text}.

\begin{figure}[h!]
	\centering
	\includegraphics{media/XapSystem.eps}
	\caption{{\bf Model of the \emph{xapAB} circuit.}
		The XapR dimers are induced by xanthosine and the induced XapR binds
		cooperatively as an activator to the \emph{xapAB} promoter. For
		these two steps, quasi-equilibrium is assumed. If both XapR binding
		sites are occupied and the polymerase is bound, the gene is
		transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
		$\gamma_{\n{m}}$, and both proteins are translated at rate
		$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
		xanthosine with Michaelis-Menten parameters $k_{a}$ and $K_a$.
		Similarly, XapB imports and exports xanthosine with parameters
		$k_{\n{b,i}}, K_{\n{b,i}}$ and $k_{\n{b,e}}, K_{\n{b,e}}$,
		respectively. Furthermore, xanthosine enters and leaves the cell
		through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
		k_{\n{nup}}$, respectively.}
	\label{fig2:model}
\end{figure}

@@

With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
probability of the promoter being in the active state, we obtain
\begin{eqnarray}
\label{eq:mbasic}
\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
\frac{
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
	e^{- \beta \Delta E_{\n{coop}}}
}{
	1 + 
	2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
}
\end{eqnarray}
Here, $w_i$ stands for the thermodynamic weight of the ith state in the
order in which they are listed in Fig~\ref{fig3:states}. As written above,
the partition function factorizes into a polymerase and a XapR term because
of our assumption of independent binding, which is further discussed
in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
p_{\n{active}}$, which we will need in the next paragraph.
In~\nameref{S1_Text}, the full chemical master equation is shown, which
leads to a \todo{MJM: I worry referees will give us grief: ``Real dists aren't Poisson,'' even though it's irrelevant}
Poisson distribution of the mRNA concentration with the same
mean.

@@

\todo[inline]{MJM: Are the following 2 paragraphs excessively detailed for main text?
maybe replace w/ 1-2 sentences about low affinity Mich-Ment transporters
that supplement XapB, \& shift all this to supplement?
If we want to emphasize the general class of model rather than the particular
biology details of Xap}
Transport of xanthosine across the cell membrane happens in two steps which
we will discuss now. All statements in this paragraph are based on data
from~\cite{Norholm2001}. In a first transport step, xanthosine passes the
outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
be a rate-limiting step. Then, XapB actively transports it across the inner
membrane, powered by the proton motive force. There are two other active
nucleoside transporters that seem to be transporting xanthosine with a very
low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
to be necessary to ``start'' the system by importing small amounts of
xanthosine that then activate \emph{xapB}, and there appears to be no other
significant way in which xanthosine can enter the cell.

We assume the kinetic scheme of all three transporters to be similar to that
of \emph{lac} permease (as it is described in~\cite{Kaback2015}). One
transporter can then do two things to change the intracellular xanthosine
concentration $\n{[x]}$: it can actively transport one xanthosine into the
cell, powered by the proton gradient, or it can transport one out of the
cell, against the proton gradient. For net transport, a proton and a
substrate need to bind on one side of the membrane and detach from the
transporter on the other side of the membrane before the transporter goes
back to the first side again (see also~\nameref{S1_Text}). Because of the
proton gradient, influx is overall thermodynamically more likely than export
at low intracellular xanthosine concentrations, which leads to a net influx.
For much higher intra- than extracellular xanthosine concentrations, the
difference in the chemical potential of xanthosine across the membrane can
dominate that of the protons and there is net efflux.

We model influx and efflux separately and use Michaelis-Menten kinetics for
XapB. For the turnover rate and Michaelis constant for influx we write
$k_{\n{b, i}}$ and $K_{\n{b, i}}$, respectively, and $k_{\n{b, e}}, K_{\n{b,
e}}$ for efflux. These parameters implicitly include the proton gradient.
For NupC and NupG, the Michaelis constant is very large, due to their low
xanthosine affinity. Thus, the general $k_{\n{cat}} \n{[E]_0}
\frac{\n{[S]}}{K_{\n{M}} + \n{[S]}}$ can be approximated as
$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \n{[S]}$. In the following,
$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0}$ is denoted by $k_{\n{nup}}$ for
influx, and the efflux rate is written as $\xi k_{\n{nup}}$. For the XapB
kinetics, no approximations can be made, because the xanthosine
concentration in the dynamic system can range from far below the respective
$K_{\n{M}}$ value to far above. The final terms for Nup and XapB are shown
below.

After transport into the cell, XapA degrades xanthosine. We model this using
standard Michaelis-Menten kinetics, with parameters $k_{\n{a}}, K_{\n{a}}$
(corresponding to turnover rate and Michaelis constant, respectively). All
the above then leads to the xanthosine rate equation
\begin{eqnarray}
\dd{\n{[x]}}{t} = \biggl(\underbrace{k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} - k_{\n{b,e}} \frac{\n{[x]}}{K_{\n{b,e}} + \n{[x]}}}_{\n{XapB}} - \underbrace{k_{\n{a}} \frac{\n{[x]}}{K_{\n{a}} + \n{[x]}}}_{\n{XapA}}\biggr) \n{[p]} + \underbrace{k_{\n{nup}} \left(c- \xi \n{[x]}\right)}_{\n{NupC\ \& \ NupG}}.
\end{eqnarray}
We recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
$c$ denotes the extracellular. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
$K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
xanthosine concentrations. At much higher intra- than extracellular
xanthosine concentrations, the efflux term takes over. More details on the
aforementioned steps and a discussion of passive diffusion can be found
in~\nameref{S1_Text}.

@@

\paragraph*{From a 3D to a 2D system.} 
Fig~\subref{fig4:bistability}{A} shows the 3D phase portrait for a
representative set of parameters (shown in Table~\ref{table1:nondim}),
whose choice is explained below. The plot looks rather complicated at first
but can be understood intuitively. The three planes are the nullclines and
the gray lines point in the direction in which the dynamical system moves at
each point. The planes intersect in three points, which are the steady-state
solutions of the dynamical system. For this choice of parameters, the system
first flows towards the mRNA nullcline, then it moves along that plane to
the intersection with the protein nullcline, and lastly, it moves along that
intersection line to one of the three intersection points of all three
planes.

\begin{figure}%[h!]
	\centering
	\includegraphics[width=1\textwidth]{media/Bistability.eps}
	\caption{{\bf Phase portraits showing bistability.}
		3D and 2D phase portraits for one set of parameters that leads to
		bistability. The parameter values are listed in
		Table~\ref{table1:nondim}. Note that all the concentrations
		($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
		$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The planes in (A) and the
		lines in (B) are the nullclines of the state variables, and their
		intersection points, marked in red in (B), are the steady-state
		solutions of the system. The region shaded in gray in (B) leads to
		negative concentrations and is unphysical. A vector plot of (B) that
		also shows the magnitude of flow at each point can be found
		in~\nameref{S1_Text}.}
	\label{fig4:bistability}
\end{figure}
%\todo{edge,size}

@@

A usual simplification with genetic circuits like this is to assume the mRNA
concentration to be at steady-state, i.e., to write $\dd{\n{[m]_a}}{\tau}=0$
and solve this for $\n{[m]_a}(\n{[p]_a},\n{[x]_a})$ to simplify the 3D to a
2D system. This restricts the dynamics to the green plane in our plot, which
is reasonable here because as explained above, the system first flows
towards that plane before either the protein or the xanthosine concentration
changes significantly. However, as already pointed out, this is different
for other parameter values, and thus, this assumption does not hold in general.
If the xanthosine dynamics are faster than the mRNA dynamics, the
system first flows towards the xanthosine nullcline. In that case, forcing
it onto the mRNA nullcline leads to significant changes in the dynamics.

@@

\paragraph*{Bistability.} 
From the 2D plot in Fig~\subref{fig4:bistability}{B}, it can clearly be seen
that for the chosen parameters, there are three steady-state solutions.
Because the 2D plane that we are in now is the mRNA nullcline, these
solutions are the same fixed points as in the 3D plot
($\dd{\n{[m]_a}}{\tau}=0$ on the nullcline and $\dd{\n{[p]_a}}{\tau}=0$,
$\dd{\n{[x]_a}}{\tau}=0$ for the 2D fixed points). One can see from the
vector field that the two outer points (labeled 1 and 3) are stable and the
middle one (labeled 2) is unstable and serves as a sort of ``switch-point''
between the other two. This means that there are two stable states the cell
can be in, one at high (point 3) and one at low (point 1) expression. As a
result, there is bistability and the distribution of expression among cells
can be bimodal, depending on initial conditions.

>>>>

	% Title must be 250 characters or less.
	\begin{flushleft}
		{\Large
			\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
		}
		\newline
		% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
		Kathrin S. Laxhuber\textsuperscript{1},
		Muir J. Morrison\textsuperscript{2},
		Rob Phillips\textsuperscript{2,3$\ast$}
		\\
		\bigskip
		\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
		\\
		\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
		\\
		\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
		\\
		\bigskip

@@

	\end{flushleft}
	% Please keep the abstract below 300 words
	\section*{Abstract}
	Membrane transporters carry important molecules through the cell membrane
	and, from a resource standpoint, should only be produced when necessary. The
	expression of membrane transporters in metabolic pathways is often
	upregulated by the transporter substrate. In \emph{E. coli}, such systems
	include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
	which encode for lactose, arabinose, and xylose transporters, respectively.
	As a case study of a minimal system, we build a generizable physical model of the \emph{xapABR} genetic
	circuit, which features a regulatory feedback loop through membrane
	transport (positive feedback) and enzymatic degradation (negative feedback)
	of an inducer. Dynamical systems analysis and stochastic simulations show
	that the membrane transport makes the model system bistable in certain
	parameter regimes. Thus, it serves as a genetic “on-off” switch, enabling
	the system to only produce a set of metabolic enzymes when the corresponding
	metabolite is present in large amounts. We find that the negative feedback
	from the degradation enzyme does not significantly disturb the positive
	feedback from the membrane transporter. We investigate hysteresis in the
	switching and discuss the role of cooperativity and multiple binding sites
	in the model circuit. Fundamentally, this work explores how a stable genetic
	switch for a set of enzymes is obtained from transcriptional auto-activation
	of a membrane transporter through its substrate.

@@

	\section*{Introduction} \todo{remove images \& todonotes everywhere for submission}
	Genetic regulatory circuits are fundamental building blocks of functioning
	cells and organisms. One abundant class of these circuits are genetic
	switches. Although their construction and function may differ,
	their common feature is bistability: their output gene expression
	will flow to and remain at one of two steady-state levels.
	The distribution of
	gene expression in a cell culture can then be bimodal. This is not to be
	confused with mere stochastic bimodality, where the system is not stable,
	and the gene expression in each cell can fluctuate between the two levels.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Circuits.eps}
		\caption{{\bf A schematic of different genetic switches.}
			(A) and (B) show the two most well-known genetic switches:
			(A) two mutual repressors and (B) a self-activating gene.
			In (C), a very much simplified version of the
			circuit that we investigate in this paper can be seen, where the
			similarity to the switch in (B) is clear. A complete version of
			the model circuit can be found in Fig~\ref{fig2:model}.}
		\label{fig1:switches}
	\end{figure}

	Through physical and mathematical modeling, we investigate a more complex
	switch system where the bistability is due, as we will show, to a
	membrane transport protein. Such a switch is common for metabolic processes in biology, for reasons discussed below. So far, models we have found are either very extensive descriptions of specific, complicated systems, or they are very basic. To enable an intuitive and fundamental understanding of the behavior, we develop a simple model of a minimal system that only has the essential components and interactions for the questions we pose, while still modeling every component explicitly and discussing the necessary model complexity for a physically correct model.

	The key feature of the type of system we investigate is the indirect activation of
	the transporter gene by the transporter substrate, leading to positive
	feedback similar to the aforementioned ``on-off'' switch. An example for
	such an architecture is the \emph{lac} operon, where lactose indirectly
	activates the expression of lactose permease. Other examples in \emph{E.
		coli} include the \emph{araFGH} and \emph{xylFGH} operons, which contain genes for
	arabinose and xylose transporters, respectively. For \emph{lac} and
	\emph{araFGH}, bistability has indeed been observed and attributed to such a
	positive feedback
	loop~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
	A eukaryotic example is the glucose transporter GLUT-2 in liver and
	\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
	complex than the following analysis.

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to accumulate. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' point and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could be involved in various cases of
	short-term adaptation.

	A key element of this mechanism is the presence of a transcription factor
	which is always expressed at a low level (often at copy numbers
	of order $\sim10$) and which binds to the transporter substrate.
	This is resource efficient for the cell, as this low copy number
	transcription factor acts as an ``always on'' sensor to detect
	the substrate, allowing high copy numbers of the membrane transporter
	and its attendant operon to be expressed only when
	their substrate is actually present. The transcription factors
	LacI, AraC, and XylR all appear to fill this
	role~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997,Song1997,Schmidt2015}.

	For our modeling, we focus on the \emph{xapABR} genetic circuit from
	\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
	complex. Instead of lactose, its purpose is to make use of the nucleoside
	xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
	circuit is made up of two operons: one that encodes for XapR and another
	that encodes for XapA and XapB. XapR appears to be a transcription factor
	that is induced by xanthosine and activates the \emph{xapAB} promoter,
	in close analogy to AraC, XylR, and also LacI, although the latter acts via repression.
	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
	but the promoter architecture and function is not yet fully understood. The
	transcription of \emph{xapR} seems to be constitutive and not
	auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
	factors suggests that XapR appears in dimers where one dimer can bind two
	xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
	nucleoside phosphorylase that degrades xanthosine into components (ribose
	and xanthine) that can be fed into metabolic
	pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
	membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine are nonspecifically transported\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it?} into and out of the cell (discussed in more detail below).
		\item XapR is induced by xanthosine.
		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
	\end{itemize}
	Because xanthosine induces the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

	\begin{figure}[h!]
		\centering
		\includegraphics{media/XapSystem.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
	rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
	probability of the promoter being in the active state, we obtain
	\begin{eqnarray}
	\label{eq:mbasic}
	\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
	p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
	\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
	\frac{
		\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
		e^{- \beta \Delta E_{\n{coop}}}
	}{
		1 + 
		2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
		\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
	}
	\end{eqnarray}
	Here, $w_i$ stands for the thermodynamic weight of the ith state in the
	order in which they are listed in Fig~\ref{fig3:states}. As written above,
	the partition function factorizes into a polymerase and a XapR term because
	of our assumption of independent binding, which is further discussed
	in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
	gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
	just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
	$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
	p_{\n{active}}$, which we will need in the next paragraph. The mean can also be found from the full chemical master equation, which is shown in~\nameref{S1_Text}.

@@

	There are two significant ways of transport of xanthosine across the cell membrane. In the induced system, the main transporter is XapB, whereas in the uninduced system, there is almost no XapB. Instead, xanthosine can enter the cell through the two nucleoside transporters NupC and NupG, which have a very low affinity for xanthosine~\cite{Norholm2001}. All these transporters, XapB, NupC, and NupG, are powered by the proton gradient across the membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme to be similar to that of the \emph{lac} permease (as it is described in~\cite{Kaback2015}). There can be import and export of xanthosine, and which one dominates depends on the proton and xanthosine concentrations on the two sides of the membrane. In both cases, a proton and a substrate need to bind to the transporter on one side of the membrane and detach from it on the other side before the empty transporter moves back to the other side again. We refer the reader to~\nameref{S1_Text} for a detailed description of the transport and its modeling, and just state the result here. We model influx and efflux separately. For XapB, we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$, $K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux. For the Nup transporters, we can linearize because of their low affinity for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for influx and $\xi k_{\n{nup}}$ for efflux.

	After transport into the cell, XapA degrades xanthosine. We model this using
	standard Michaelis-Menten kinetics, with parameters $k_{\n{a}}, K_{\n{a}}$
	(corresponding to turnover rate and Michaelis constant, respectively). Transport and degradation then leads to the xanthosine rate equation
	\begin{eqnarray}
	\dd{\n{[x]}}{t} = \biggl(\underbrace{k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} - k_{\n{b,e}} \frac{\n{[x]}}{K_{\n{b,e}} + \n{[x]}}}_{\n{XapB}} - \underbrace{k_{\n{a}} \frac{\n{[x]}}{K_{\n{a}} + \n{[x]}}}_{\n{XapA}}\biggr) \n{[p]} + \underbrace{k_{\n{nup}} \left(c- \xi \n{[x]}\right)}_{\n{NupC\ \& \ NupG}}.
	\end{eqnarray}
	Recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
	$c$ denotes the extracellular. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
	$K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
	xanthosine concentrations. At much higher intra- than extracellular
	xanthosine concentrations, the efflux term takes over. More details on the
	aforementioned steps and a discussion of passive diffusion can be found
	in~\nameref{S1_Text}.

@@

	\paragraph*{From a 3D to a 2D system.} 
	Fig~\subref{fig4:bistability}{A} shows the 3D phase portrait for a
	representative set of parameters (shown in Table~\ref{table1:nondim}),
	whose choice is explained below. The plot looks rather complicated at first
	but can be understood intuitively. The three surfaces are the nullcline surfaces and
	the gray lines point in the direction in which the dynamical system moves at
	each point. The surfaces intersect in three points, which are the steady-state
	solutions of the dynamical system. For this choice of parameters, the system
	first flows towards the mRNA nullcline, then it moves along that surface to
	the intersection with the protein nullcline, and lastly, it moves along that
	intersection line to one of the three intersection points of all three
	surfaces.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Bistability.eps}
		\caption{{\bf Phase portraits showing bistability.}
			3D and 2D phase portraits for one set of parameters that leads to
			bistability. The parameter values are listed in
			Table~\ref{table1:nondim}. Note that all the concentrations
			($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
			lines in (B) are the nullclines of the state variables, and their
			intersection points, marked in red in (B), are the steady-state
			solutions of the system. The region shaded in gray in (B) leads to
			negative concentrations and is unphysical. A vector plot of (B) that
			also shows the magnitude of flow at each point can be found
			in~\nameref{S1_Text}.}
		\label{fig4:bistability}
	\end{figure}

@@

	A usual simplification with genetic circuits like this is to assume the mRNA
	concentration to be at steady-state, i.e., to write $\dd{\n{[m]_a}}{\tau}=0$
	and solve this for $\n{[m]_a}(\n{[p]_a},\n{[x]_a})$ to simplify the 3D to a
	2D system. This restricts the dynamics to the green surface in our plot, which
	is reasonable here because as explained above, the system first flows
	towards that surface before either the protein or the xanthosine concentration
	changes significantly. However, as already pointed out, this is different
	for other parameter values, and thus, this assumption does not hold in general.
	If the xanthosine dynamics are faster than the mRNA dynamics, the
	system first flows towards the xanthosine nullcline. In that case, forcing
	it onto the mRNA nullcline leads to significant changes in the dynamics.

@@

	\paragraph*{Bistability.} 
	We map the mRNA nullcline surface (green in Fig~\subref{fig4:bistability}{A}) onto a plane to show it as the 2D plot in Fig~\subref{fig4:bistability}{B}.
	From this 2D plot, it can clearly be seen
	that for the chosen parameters, there are three steady-state solutions.
	Because the system is restricted to the mRNA nullcline surface, these steady-state solutions are the same as those in the 3D plot
	($\dd{\n{[m]_a}}{\tau}=0$ on the nullcline and $\dd{\n{[p]_a}}{\tau}=0$,
	$\dd{\n{[x]_a}}{\tau}=0$ for the 2D fixed points). One can see from the
	vector field that the two outer fixed points (labeled 1 and 3) are stable and the
	middle one (labeled 2) is unstable and serves as a sort of ``switch-point''
	between the other two. This means that there are two stable states the cell
	can be in, one at high (point 3) and one at low (point 1) expression. As a
	result, there is bistability and the distribution of expression among cells
	can be bimodal, depending on initial conditions.
